---
title: "Water Quality and Potability"
output: html_notebook
--- 


We are collecting a dataset on water quality to train a machine learning model
for binary classification: determining whether water is safe for consumption (1)
or not (0). This model will help with water treatment decisions and ensure
compliance with quality standards.
We applied different summarization and plotting methods to help us to understand
our dataset, such as scatter, histogram and bar plot. Then, we applyed
preprocess in our data using data cleaning, data transformation 
and feature selection.

```{r}
#library:
#install.packages("caret")
#install.packages("glmnet")
#install.packages("Boruta")
#install.packages("mlbench")
#install.packages("randomForest")
library(outliers)
library(dplyr)
library(mlbench)
library(caret)
library(glmnet)
library(Boruta)
library(ggplot2)
library(randomForest)

getwd()
#setwd("/Users/mahayie/Desktop/326p")
#getwd()

water_potability = read.csv('Dataset/water_potability.csv')

View(water_potability)

str(water_potability)
summary(water_potability)
```

Checking for missing values:
```{r}
dim(water_potability)
sum(is.na(water_potability))
```
Remove rows with missing values
```{r}
water_potability = na.omit(water_potability)
View(water_potability)
```

Description:
The absence of data in certain variables or columns in a dataset is referred to as missing or null values due to various reasons.  It can have a negative impact on the dataset's efficiency and the information that can be taken from it later, so we checked to see whether our data had missing or null values and eliminated these rows to produce a more efficient dataset.


Standard deviation:
```{r}
sd(water_potability$Turbidity)
sd(water_potability$Solids)
sd(water_potability$Conductivity)
sd(water_potability$Organic_carbon)
sd(water_potability$ph)
```

Mean:
```{r}
mean(water_potability$Turbidity)
mean(water_potability$Solids) 
mean(water_potability$Conductivity) 
mean(water_potability$Organic_carbon) 
mean(water_potability$ph) 
```

Median
```{r}
median(water_potability$Turbidity)
median(water_potability$Solids)
median(water_potability$Conductivity)
median(water_potability$Organic_carbon)
median(water_potability$ph)
```

Variance
```{r}
var(water_potability$Turbidity)
var(water_potability$Solids)
var(water_potability$Conductivity)
var(water_potability$Organic_carbon)
var(water_potability$ph)
```

Statistical Measures:
```{r}
summary(water_potability$Conductivity)
summary(water_potability$Organic_carbon)
summary(water_potability$Hardness)
summary(water_potability$Solids)
summary(water_potability$Chloramines)
summary(water_potability$Potability)
summary(water_potability$Sulfate)
summary(water_potability$Trihalomethanes)
summary(water_potability$Turbidity)
summary(water_potability$ph)
```

Descriotion:
With using minimum, maximum, mean, median laws it helps to provide an overview of the data's key characteristics


outliers
before removing outlier:
```{r}
dim(water_potability)
head(water_potability)
```

removing outliers:

- ph
```{r}
summary(water_potability$ph)
quartiles <- quantile(water_potability$ph, probs = c(.25, .75), na.rm = FALSE)
quartiles
iqr <- IQR(water_potability$ph)
iqr
lower <- quartiles[1] - 1.5*iqr
lower
upper <- quartiles[2] + 1.5*iqr
upper

boxplot(ph ~ Potability, data = water_potability)

repeat {
  out_val <- boxplot(water_potability$ph, ylab = 'ph')$out
  out_val
  out_rows <- which(water_potability$ph %in% c(out_val))
  out_rows
  
  if(sum(out_rows) > 0) water_potability <- water_potability[-out_rows,]
  else {break}
}
summary(water_potability$ph)

#-------------------------------------------
```

-Hardness
```{r}
summary(water_potability$Hardness)
quartiles <- quantile(water_potability$Hardness, probs = c(.25, .75), na.rm = FALSE)
quartiles
iqr <- IQR(water_potability$Hardness)
iqr
lower <- quartiles[1] - 1.5*iqr
lower
upper <- quartiles[2] + 1.5*iqr
upper

boxplot(Hardness ~ Potability, data = water_potability)

repeat {
  out_val <- boxplot(water_potability$Hardness, ylab = 'Hardness')$out
  out_val
  out_rows <- which(water_potability$Hardness %in% c(out_val))
  out_rows

  if(sum(out_rows) > 0) water_potability <- water_potability[-out_rows,]
  else {break}
}
summary(water_potability$Hardness)

#-------------------------------------------
```

-Solids
```{r}
summary(water_potability$Solids)
quartiles <- quantile(water_potability$Solids, probs = c(.25, .75), na.rm = FALSE)
quartiles
iqr <- IQR(water_potability$Solids)
iqr
lower <- quartiles[1] - 1.5*iqr
lower
upper <- quartiles[2] + 1.5*iqr
upper

boxplot(Solids ~ Potability, data = water_potability)

repeat {
  out_val <- boxplot(water_potability$Solids, ylab = 'Solids')$out
  out_val
  out_rows <- which(water_potability$Solids %in% c(out_val))
  out_rows

  if(sum(out_rows) > 0) water_potability <- water_potability[-out_rows,]
  else {break}
}
summary(water_potability$Solids)

#-------------------------------------------
```

-Chloramines
```{r}
summary(water_potability$Chloramines)
quartiles <- quantile(water_potability$Chloramines, probs = c(.25, .75), na.rm = FALSE)
quartiles
iqr <- IQR(water_potability$Chloramines)
iqr
lower <- quartiles[1] - 1.5*iqr
lower
upper <- quartiles[2] + 1.5*iqr
upper

boxplot(Chloramines ~ Potability, data = water_potability)

repeat {
  out_val <- boxplot(water_potability$Chloramines, ylab = 'Chloramines')$out
  out_val
  out_rows <- which(water_potability$Chloramines %in% c(out_val))
  out_rows

  if(sum(out_rows) > 0) water_potability <- water_potability[-out_rows,]
  else {break}
}
summary(water_potability$Chloramines)

#-------------------------------------------

```

-Sulfate
```{r}
summary(water_potability$Sulfate)
quartiles <- quantile(water_potability$Sulfate, probs = c(.25, .75), na.rm = FALSE)
quartiles
iqr <- IQR(water_potability$Sulfate)
iqr
lower <- quartiles[1] - 1.5*iqr
lower
upper <- quartiles[2] + 1.5*iqr
upper

boxplot(Sulfate ~ Potability, data = water_potability)

repeat {
  out_val <- boxplot(water_potability$Sulfate, ylab = 'Sulfate')$out
  out_val
  out_rows <- which(water_potability$Sulfate %in% c(out_val))
  out_rows

  if(sum(out_rows) > 0) water_potability <- water_potability[-out_rows,]
  else {break}
}
summary(water_potability$Sulfate)

#-------------------------------------------

```

-Conductivity
```{r}
summary(water_potability$Conductivity)
quartiles <- quantile(water_potability$Conductivity, probs = c(.25, .75), na.rm = FALSE)
quartiles
iqr <- IQR(water_potability$Conductivity)
iqr
lower <- quartiles[1] - 1.5*iqr
lower
upper <- quartiles[2] + 1.5*iqr
upper

boxplot(Conductivity ~ Potability, data = water_potability)

repeat {
  out_val <- boxplot(water_potability$Conductivity, ylab = 'Conductivity')$out
  out_val
  out_rows <- which(water_potability$Conductivity %in% c(out_val))
  out_rows

  if(sum(out_rows) > 0) water_potability <- water_potability[-out_rows,]
  else {break}
}
summary(water_potability$Conductivity)

#-------------------------------------------
```

-Organic_carbon
```{r}
summary(water_potability$Organic_carbon)
quartiles <- quantile(water_potability$Organic_carbon, probs = c(.25, .75), na.rm = FALSE)
quartiles
iqr <- IQR(water_potability$Organic_carbon)
iqr
lower <- quartiles[1] - 1.5*iqr
lower
upper <- quartiles[2] + 1.5*iqr
upper

boxplot(Organic_carbon ~ Potability, data = water_potability)

repeat {
  out_val <- boxplot(water_potability$Organic_carbon, ylab = 'Organic_carbon')$out
  out_val
  out_rows <- which(water_potability$Organic_carbon %in% c(out_val))
  out_rows

  if(sum(out_rows) > 0) water_potability <- water_potability[-out_rows,]
  else {break}
}
summary(water_potability$Organic_carbon)

#-------------------------------------------
```

-Trihalomethanes
```{r}
summary(water_potability$Trihalomethanes)
quartiles <- quantile(water_potability$Trihalomethanes, probs = c(.25, .75), na.rm = FALSE)
quartiles
iqr <- IQR(water_potability$Trihalomethanes)
iqr
lower <- quartiles[1] - 1.5*iqr
lower
upper <- quartiles[2] + 1.5*iqr
upper

boxplot(Trihalomethanes ~ Potability, data = water_potability)

repeat {
  out_val <- boxplot(water_potability$Trihalomethanes, ylab = 'Trihalomethanes')$out
  out_val
  out_rows <- which(water_potability$Trihalomethanes %in% c(out_val))
  out_rows

  if(sum(out_rows) > 0) water_potability <- water_potability[-out_rows,]
  else {break}
}
summary(water_potability$Trihalomethanes)

#-------------------------------------------
```

-Turbidity
```{r}
summary(water_potability$Turbidity)
quartiles <- quantile(water_potability$Turbidity, probs = c(.25, .75), na.rm = FALSE)
quartiles
iqr <- IQR(water_potability$Turbidity)
iqr
lower <- quartiles[1] - 1.5*iqr
lower
upper <- quartiles[2] + 1.5*iqr
upper

boxplot(Turbidity ~ Potability, data = water_potability)

repeat {
  out_val <- boxplot(water_potability$Turbidity, ylab = 'Turbidity')$out
  out_val
  out_rows <- which(water_potability$Turbidity %in% c(out_val))
  out_rows

  if(sum(out_rows) > 0) water_potability <- water_potability[-out_rows,]
  else {break}
}
summary(water_potability$Turbidity)
```


After removing outliers:
```{r}
dim(water_potability)
str(water_potability)
head(water_potability)
```

Description:
Removing outliers from a dataset is critical for assuring the quality and reliability of statistical analysis and machine learning models. We found all outliers in the numerical attributes and subsequently eliminated the rows containing the outliers. 



Charts

Histogram
```{r}
hist(water_potability$ph)
hist(water_potability$Chloramines)
hist(water_potability$Hardness)
```

Bar Plot
```{r}
water_potability$Potability %>% table() %>% barplot()

# add colors
bb <- water_potability$ph %>% table() %>% barplot( main='ph',col=c('pink'))

```

Pie chart
```{r}
water_potability$Potability %>% table() %>% pie() # plot pie chart without percentages

tab <- water_potability$Potability %>% table()
precentages <- tab %>% prop.table() %>% round(3)*100
txt <- paste0(names(c("Not Potable","Potable")),'\n',precentages,'%')
pie(tab,labels = c("Not potabale","Potabale"))
```

Scatter Plot
```{r}
with(water_potability, plot(Turbidity, ph, col = Potability, pch = as.numeric(Potability)))
```
 
Description:
-Histogram:
  The histogram shows the frequency of ph in the dataset; we noted that the majority of values fall    within the usual range, which is about between 6 and 8, but it also shows several outliers.
-Scatter plot:
  This scatter demonstrates the correlation and proportionality between the two qualities, allowing    us to establish whether or not turbidity and pH are connected.
-Bar Plot
  the bar plot represent how ph levels affect water portability in the dataset it indicates that ph    level above 10 is not portibal and humans cant consume it 



Remove Redundant Features:
```{r}
correlation_matrix <- cor(water_potability[,1:9])
high_correlation_features <- findCorrelation(correlation_matrix, cutoff = 0.5)
print(high_correlation_features)
heatmap(correlation_matrix)
```
Description:
This will find the correlation between the features and represent it in heat map



Feature selection

Rank Features By Importance:
```{r, warning=FALSE}
#train random forest model and calculate feature importance
rf = randomForest(x= water_potability[,1:9],y= water_potability[,10])
var_imp <- varImp(rf, scale = FALSE)
#sort the score in decreasing order
var_imp_df <- data.frame(cbind(variable = rownames(var_imp), score = var_imp[,1]))
var_imp_df$score <- as.double(var_imp_df$score)
var_imp_df[order(var_imp_df$score,decreasing = TRUE),]

ggplot(var_imp_df, aes(x=reorder(variable, score), y=score)) + 
  geom_point() +
  geom_segment(aes(x=variable,xend=variable,y=0,yend=score)) +
  ylab("IncNodePurity") +
  xlab("Variable Name") +
  coord_flip()
```

Recursive Feature elimination:
```{r, warning=FALSE}
control <- rfeControl(functions=rfFuncs, method="cv",number=10)
rf <- trainControl(method = "cv", number = 10, verboseIter = FALSE)
# run the RFE algorithm
rfe_model <- rfe(x= water_potability[,1:9],y= water_potability[,10], sizes=c(1:9), rfeControl=control)
# summarize the results
print(rfe_model)
# list the chosen features
predictors(rfe_model)
# plot the results
plot(rfe_model, type=c("g", "o"))
```
Description:
ranking features by importance is a technique used to identify the most influential variables in a dataset for predicting a target variable. This process helps in understanding which features have the most impact on the model's performance. By ranking features by importance.

removing redundant features refers to the process of eliminating variables or features from a dataset that do not provide any additional or unique information.



Data transformation

Normlization
```{r}
normalize=function(x){return ((x-min(x))/(max(x)))}
w = water_potability

water_potability$Sulfate=normalize(water_potability$Sulfate)
water_potability$Conductivity=normalize(water_potability$Conductivity)
water_potability$Turbidity=normalize(water_potability$Turbidity)
water_potability$ph=normalize(water_potability$ph)
water_potability$Chloramines=normalize(water_potability$Chloramines)
water_potability$Solids=normalize(water_potability$Solids)
water_potability$Trihalomethanes=normalize(water_potability$Trihalomethanes)
water_potability$Organic_carbon=normalize(water_potability$Organic_carbon)
water_potability$Hardness=water_potability$Hardness/1000

water_potability$Hardness<-normalize(water_potability$Hardness)
print(water_potability)
```
Description:
Normalization refers to the process of scaling variables to have a common range. It helps in comparing variables with different scales. In the solids attribute will create critical challenges since of the huge and diverted values(min=320,9 max=43195.5) so we normalized the solids to make values smaller and more reasonable. Also we normalized all the scaled attributes:Sulfate,Conductivity,Organic_corbon, Trihalomerhanes, Turbidity,ph,Chloramines.



Discretization:
```{r}
w$Trihalomethanes= cut(w$Trihalomethanes, breaks = seq(0,125,by=25),right=FALSE)
w$Solids= cut(w$Solids, breaks = seq(0,50000,by=10000),right=FALSE)
w$Organic_carbon= cut(w$Organic_carbon, breaks = seq(0,25,by=5),right=FALSE)
print(w)
```
Description:
Discretization is the process of transforming continuous variables into discrete or categorical variables. It’s can be useful for analyzing data that has a large number of unique values or when you want to simplify the data.So In Trihalomethanes we intervals by dividing the values by 25 to have a labels with equal width : (0,25],(25,50],(50,75],(75,100],(100,125].


Encoding
encoding is the process of converting characters or strings into a specific encoding format. Since we don’t have a Nominal  attribute in our database we couldn’t implement it.




